Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-09 demystify.tex
updates

The primary focus of the admissions committee is to \textbf{evaluate your background and interest in research} (a PhD in Computer Science is a research degree!), and determine if you would \textbf{fit into} the program. To evaluate your profile, we consider
the following key indicators.

@@

Tenured faculty, e.g., associate and full profs., are likely older, more well-known, and have more experience in managing students.  However, they might not push you as hard. They might also expect you to figure things out yourself, i.e., so you need to be independent.  Some tenured faculty are also no longer active in research and more involved with administrative responsibilities or with their startup companies (this means they will likely not take new students).

@@

In many cases you do not need to do a PhD in CS to work in your area of interested. For example, in addition to a traditional CS department, GMU has IST and Cybersecurity departments, which have faculty  with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  So it is totally possible that you still get to do CS research and publish in CS-focused venues even if you're not in a traditional CS program.  It is  common to see faculty with CS PhD in a non-CS department as well as faculty with non-CS PhD in CS department.

However, if your goal is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. A non-CS may \emph{co-advise} or be in the PhD dissertation committee, but your main PhD adviser will likely be a tenured or tenure-track faculty in CS (\autoref{sec:faculty-types}). If in doubt, check with the CS department for their requirements.

@@

Most of what is written in this document is applicable to both domestic\footnote{Domestic simply means you did your undergrad (or MS) at a US university.} and international students.  However, there are some differences and benefits that domestic students should be aware of.  This section lists some of these.

@@

\paragraph{School \autoref{sec:your-school}} we already know about your school, which is a plus (see \autoref{sec:your-school}). You are also more familiar with the US education system and culture (so hopefully you are aware of the cultural issues listed in \autoref{sec:cultural}).

@@

>>>>

The focus of the admissions committee is to \textbf{evaluate your background and interest in research} to see if you can handle a PhD in CS, which is a \emph{research} degree!. We also aim to determine if you would \textbf{fit into the program}. To evaluate your profile, we consider
the following key indicators.

@@

Tenured faculty, e.g., associate and full profs., are likely older, more well-known, and have more experience in managing students.  However, they might not push you as hard. They might also expect you to figure things out yourself, i.e., you need to be independent.  Some tenured faculty are also no longer active in research and more involved with administrative responsibilities or with their startup companies (this means they will likely not take new students).

@@

In many cases you do not need to do a PhD in CS to work in your area of interest. For example, in addition to a traditional CS department, GMU has IST and Cybersecurity departments, both of which have faculty  with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  So it is totally possible that you still get to do CS research and publish in CS-focused venues even if you're not in a traditional CS program.  It is  common to see faculty with CS PhD in a non-CS department as well as faculty with non-CS PhD in CS department.

However, if your goal is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. A non-CS faculty may \emph{co-advise} or be in the PhD dissertation committee, but your main PhD adviser will likely be a tenured or tenure-track faculty in CS (\autoref{sec:faculty-types}). If in doubt, check with the CS department for their requirements.

@@

Most of what is written in this document is applicable to both domestic\footnote{Here domestic means you did your undergrad (or MS) at a US university.} and international students.  However, there are some differences and benefits that domestic students should be aware of.  This section lists some of these.

@@

\paragraph{School \autoref{sec:your-school}} We already know about your school, which is a plus (see \autoref{sec:your-school}). You are also more familiar with the US education system and culture (so hopefully you are aware of the cultural issues listed in \autoref{sec:cultural}).

@@

\chapter{Research Opportunities for Undergraduates}\label{sec:undergrad-research}

As discussed in \autoref{sec:undergrad-research}, research experience is a crucial component of a strong PhD application. It gives you opportunities to try out research, determine what research area you're interested in, publish papers (\autoref{sec:research-ability}), and get strong LoRs (\autoref{sec:lor}).

 This section provides some guidance on how to gain research experience as an undergraduate student.  It also gives advice to international students or domestic ones from small colleges where research opportunities are limited.

\paragraph{Locally} The most common way to find research is right at your own institution.  
If you did well or like a class, you can check with the professor of that class for research opportunities.  
You can also go to the department directory and then professors' websites and see if they are looking for undergraduate researchers.
Even if they say they are looking for graduate students, you still should contact them and ask.

You can also take honor thesis or independent study courses with a professor.  This is a good way to get research experience and also get a LoR from the professor.  You can also ask your academic adviser or other faculty members for suggestions.  Finally, you can also ask your peers who are already doing research.  You can use the method described in \autoref{sec:contact} to contact professors

\begin{commentbox}[Vu:] 
    I enjoy working with undergrads and have advised many (10+).
    I often recruit undergraduate researchers through my classes, e.g., asking students who did well in my class if they are interested in research.  Occasionally I was introduced to students by other students or faculty.

    AT GMU we pay undergrads about \$15/hr and about 10 hrs/week. This is a good way to make some money and get research experience.
  \end{commentbox}

% \paragraph{Online Research Communities and Open Source Contributions} Places such as Github offer many opportunities for research, even though they might not appear so (e.g., they do not explicitly mention research). 
% In many cases, projects are directly related to research (e.g., open-source software from professors' or research labs). For example, my research group has a many open-source projects that undergrads can contribute to (see \url{https://github.com/dynaroars/}).

%  By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field. Not only you gain research experience, you might be able to get a LoR from the project maintainer and definitely should write about this experience in your SOP (\autoref{sec:research-statement}.)

% \paragraph{Virtual Research Opportunities Beyond Physical Boundaries}: Several universities, organizations, and research institutions offer virtual internships and research programs aimed at providing hands-on research experience. These programs often involve working remotely under the guidance of experienced mentors and collaborating with a team of fellow researchers. For instance, \href{https://docs.google.com/forms/d/1btIwt4HwjyKMOUk-EMy3rbkfWzFxv2lNrMm_zkd0pA4/viewform?edit_requested=true}{UIUC+ Summer Undergraduate Research in Software Engineering}  offers an unpaid remote internship for software engineering students all over the world to collaborate with mentors from University of Illinois at Urbana Champaign.

% \paragraph{Online Conferences and Workshops}: Attending online conferences and workshops is another way to gain exposure to cutting-edge research and establish connections with experts in the field. Many conferences now provide virtual participation options, enabling students to access research talks, poster sessions, and panel discussions and sometimes access designated chat rooms or networking events where participants can engage with researchers, ask questions, and seek potential research collaborations. It is beneficial to also create profiles academic collaboration platforms such as ResearchGate. By creating profiles, joining relevant research groups, and participating in discussions, students can connect with established researchers, contribute to ongoing projects, and potentially collaborate on publications or research proposals as they provide access
